\LARGE \bf Stochastic Uncertainty Propagation in Power System Dynamics using Measure-valued Proximal Recursions

### 2022-03-09 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

\noindent
{\bf AE.1. Comments by AE}:\\
{\em The authors are requested to revise and resubmit their paper after addressing reviewer feedback. Of particular importance is that reviewers point out the model leveraged for analysis is overly simplistic.}

{\nib {\blue We are indebted to the AE and the reviewers for their help to improve the manuscript.

In addition to the itemized responses to the reviewers' comments and the corresponding revisions marked in blue in the manuscript, we have made the \ul{following additional fixes and improvements} (also marked in blue in the revised manuscript):\\
(i) In equation (8), we corrected a typo in its right hand side: the second summand runs over pairs $(i,j)$ such that $i<j$.\\
(ii) To make the presentation tighter, the Appendix appearing in the previous version has been eschewed. We have accordingly adjusted and condensed the material in Sec. III-D of the revised manuscript.\\
(iii) Some mathematical typos have been fixed in the revised manuscript's Sec. III-C and III-D.

Regarding the usage of simple model, please see our response to comment \textbf{R3.1} and \textbf{1.3, sub-comment (1)}. In summary, the current state-of-the-art in power system literature for this problem is ref. [45] with second order dynamics + single machine infinite bus ($n=1$ in our manuscript's notation). Given this context, our manuscript enables a new kind of computation that can handle moderately large $n$ with second order sample path dynamics. This computation is online and approximates neither the statistics nor the dynamics. We believe the technical contribution of this paper is a necessary stepping stone to handle more complex models in follow up research. 
}}

@@

\ul{Following the comment \textbf{R4.3} to improve the presentation, in the revised manuscript, we have expanded the last two paragraphs of Sec. III-A with a new reference [50]}. Please see our response to that comment below for details.

\ul{As detailed in the response to comment \textbf{R3.4}, in the revised manuscript, we have enlarged Figures 5 and 6, and expanded the plot in Fig. 6}. There we also explained why those plot styles are used.

\ul{Furthermore, Fig. 7 in the revised manuscript has also been expanded to include the univariate angular velocity transient marginal plots for all the five generators of the IEEE 14 bus system; previous version had this plot only for the generator 4 (buss 6)}.

{In addition to the above, \ul{we have included the following new figures in the revised manuscript}: Figs. 8, 10, 13, 14. In particular, Fig. 8 addresses comment \textbf{R2.5}. Fig. 10 addresses comment \textbf{R1.7}. Figs. 13 and 14 address comments \textbf{R1.4, R4.2 and R4.7}.
}}

@@

\textbf{Regarding OU process in load:} \ul{The reviewer might have missed the following key point of our paper}. The Kron reduced second order stochastic model implies that in the unreduced sense, the load dynamics was first order noisy Kuramoto, in addition to the synchronous machine dynamics being second order noisy Kuramoto. Therefore, we do have stochastic load dynamics, which is in fact more realistic than OU process model. While OU process assumes linear drift with constant diffusion, whereas our setting considers the load SDEs having drifts with trigonometric nonlinearity and constant diffusion.

}}

\noindent
{\bf R1.4. Comments by reviewer}:\\
{\em There is no comparison of the proposed method compared with other state-of-the-art methods. Therefore, it is hard for the readers to judge the importance of this article.}

{\nib{ \blue \ul{Regarding the comparison with existing methods}, we have provided comparison with Monte Carlo in the revised manuscript's Figs. 12, 13 and 14; \ul{among these Figs. 13 and 14 are new} (in response to comment \textbf{R4.7}). Also, the \ul{new Fig. 8 in the revised manuscript shows the consistency for the rotor angle stochastic predictions w.r.t. Monte Carlo for the simulation reported in Sec. V-A}, as requested by comment \textbf{R2.5}. Please also see our response to comment \textbf{R4.2} where we explain why propagating the joint PDF for even a modest 5 generator system with finite element simulation is challenging with general purpose computers.}}

@@

\noindent
{\bf R1.5. Comments by reviewer}:\\
{\em Contributions are over-claimed in this article:\\
(1) The scalability of the proposed method cannot convince the readers since you are using a too simple model and ignored many important random dimensions from the loads.\\
(2) The proposed method claims to be nonparametric, however, all the inputs are assumed to be either wiener process or uniform distribution, which are all parametric distributions.\\
(3) If the ``nonparametric" is only referred to the output PDF, I would not think of it as a major contribution since there are a lot of existing techniques that can depict nonparametric PDF.}

{\nib{ \blue \ul{Our responses to the sub-comments below}.

(1) \textbf{Regarding simple model and ignoring the randomness in loads}: Please refer to our response to \textbf{R1.3, sub-comment (1)} where we explained why this model is relevant for this particular problem, and why this is not limiting to our methodological contribution. In response to \textbf{R1.3, last unnumbered sub-comment in bold text: ``Regarding OU process in load"}, we explained that we are not ignoring randomness from loads.

(2) \textbf{Regarding nonparametric}: \ul{We are \emph{not} making any parametric assumptions}. Our usage of the word ``nonparametric" is standard as in mathematical statistics literature. That the additive noise process is Wiener does not invite any loss of generality since if it was colored, we could always re-cast the same in our SDE form via a linear map. \ul{Furthermore, we have zero assumptions on the initial joint state PDF $\rho_{0}$ and on the joint PDF of the parameters appearing in the model. As a matter of fact, we don't even need to know the functional form of these joint PDFs. For instance, our framework remains unchanged if we are given the initial joint state and parameter distributional data as numerical table containing samples from these joints \emph{as well as} the joint PDF values evaluated at those samples}. The latter might be coming from some measurement-based sensor fusion/Bayesian algorithm. The specific product von Mises and uniform PDFs in Sec. V-A and V-B are only used for generating example data to perform our illustrative numerical simulations. %\ul{This again underscores how looking into numerical simulations without contemplative reading and understanding of the earlier Sections may lead to misleading conclusions}.

@@

(3) \textbf{Regarding depict nonparametric PDF}: \ul{This paper is not about nonparametric \emph{depiction} but about nonparametric \emph{computation} of the transient joint state PDFs}. For example, Monte Carlo followed by function approximation via Kernel density estimates and its variants, or other function approximants (e.g., RBF, Gaussian mixture, neural networks) are all parametric methods. We enable computing the value of joint state PDF at a time-varying sample location \emph{online (not as offline post-processing)} without assuming any functional representation of the transient joint which the computed values correspond to. Notice that we are doing this joint computation without gridding the state space, which would be impossible had we followed the traditional scientific computing route: finite element/volume methods, tensor decomposition techniques etc. \ul{We believe the proposed computational framework is novel, highly non-trivial and is specifically useful in this power system uncertainty propagation context}.
}}

\noindent
{\bf R1.6. Comments by reviewer}:\\
{\em The simulation time is too short to convince the readers. Only 1 second is considered here. Note that the errors can be quickly accumulated in the iteration.}

{\nib{ \blue Please note that our joint PDF computation has no truncation error accumulation issue (as in for example, moment based propagation techniques where such errors arise from a priori assuming finite dimensionality of the sufficient statistic). Also, the new Fig. 10 reports the total computational time for 1 minute time horizon. Please also see the additional error statistics included in the revised version's Fig. 13 and Fig. 14.}}

@@

\noindent
{\bf R1.7. Comments by reviewer}:\\
{\em It is suggested to count the total computing time, not on each step.}

{\nib{ \blue Thanks. \ul{We have included new Fig. 10 which reports the total computational time for 1 minute time horizon} for all the three simulation cases in Sec. V-A and V-B. For each of these cases, Fig. 10 shows the total computational time across 100 simulation instances. This Fig. 10 is referred to in text in Sec. V-A (last paragraph), and also in Sec. V-B (third paragraph, last sentence).

While showing the per-step computational times in Figs. 9 and 11 in the revised manuscript are somewhat unconventional, in the revised manuscript's Sec. V-A end, \ul{we have added detailed explanation (highlighted in blue)} of why such depictions are meaningful in our computation.}}

@@

{\nib \blue{Thanks for this comment. In the previous version, the strict 10 page limit enforced small figure sizes. In the revised manuscript, \ul{Figures 5 and 6 are made larger}. Also, the Fig. 6 in the revised manuscript shows the transient $(\theta_i,\omega_i)$ bivariate marginals for \ul{all the five generators in Case II}; the previous version showed the same only for the generator 4 (bus 6). We believe these changes have improved clarity.

Regarding the style of plots, we had considered plotting the bivariate marginals as planar contour plots (as in Fig. 1 in this rebuttal document) \ul{but decided against it because of the following two reasons}.

\emph{First}, \ul{in our opinion, this kind of plot is scientifically misleading as they hide the fact that these bivariate marginal PDFs are not really supported on the plane, or the subsets thereof, but instead have cylindrical supports}. Our message that the proposed method explicitly accounts this manifold geometry in the computation, would be missed had we chosen to make the plots the other way.%as per populist cognitive bias on behalf of the readers. That the planar plots evoke a sense of familiarity cannot be a scientific argument in favor of it.

\emph{Second}, \ul{the planar contour marginal plots require additional mental efforts from the readers to correctly parse the statistical information conveyed}. For instance, consider the fourth column, second row subplot in Fig. 1 in this rebuttal document. It is upon the reader to parse that the two high probability regions on the left and right of that subplot are actually referring to the same angular high probability region around $\theta_{2}=0$ rad, thanks to the periodic boundary condition on the angular variable. %Our philosophy of technical communication is that the onus is on the authors, not on the readers, to clarify the scientific information as correctly as possible. %We are aware that many papers in the power system literature make plots in the rectangular domain ignoring the manifold nature of the support, which in our experience, continue to add pain to the graduate students and beginning researchers trying to decipher the statistical meaning conveyed in those plots. 
We thus prefer to make plots that depict the variables as they are.}}

@@

\noindent
{\bf R4.1. Comments by reviewer}:\\
{\em The main contribution of this paper lies in applying results on the governing partial differential equation (PDE) for the joint PDF to avoid discretization of the state space, thereby circumventing computational challenges. The reviewer appreciates that this paper is rigorous in development and thinks it deals with an important problem.}

{\nib \blue{We appreciate the positive remark.}}

\noindent
{\bf R4.2. Comments by reviewer}:\\
{\em The key contribution of this paper is that the proposed method can reduce the computation burden when solving for the transient PDF. However, no comparison is provided with existing methods. The authors should provide comparisons to the exiting approaches, for example, based on finite element discretization or other numerical methods. The test example could be performed using the 5-generator system or larger systems. Besides, referring to Figs. 8 and 9, the total computation time should be reported for a longer horizon.}

{\nib \blue{\ul{Regarding the comparison with existing methods}, we have provided comparison with Monte Carlo in the revised manuscript's Figs. 12, 13 and 14; \ul{among these Figs. 13 and 14 are new} (in response to comment \textbf{R4.7}). Also, the \ul{new Fig. 8 in the revised manuscript shows the consistency for the rotor angle stochastic predictions w.r.t. Monte Carlo for the simulation reported in Sec. V-A}, as requested by comment \textbf{R2.5}.

Please note that performing finite element simulation for 5-generator system is a very tall order since the state space then is 10 dimensional, and even if we do a very coarse 10 discretizations per dimension, then such a simulation would require us to evolve a joint PDF over $10^{10}$ grid points. This requires dedicated high-performance computing architecture and an undertaking well beyond the scope of this paper. This also explains why the state-of-the-art for direct computation of transient joint PDF in power systems literature remains single-machine-infinite-bus $(n=1)$ as in the 2013 ref. [45] which indeed performed the finite element simulation with a two dimensional discretization, and mentioned the difficulty of scaling this computational approach for even moderately large sized systems.

\ul{Regarding the computational time for longer horizon}, we have included new Fig. 10 which reports the total computational time for 1 minute time horizon for all the three simulation cases in Sec. V-A and V-B. For each of these cases, Fig. 10 shows the total computational time across 100 simulation instances. This Fig. 10 is referred to in text in Sec. V-A (last paragraph), and also in Sec. V-B (third paragraph, last sentence).}}

@@

\noindent
{\bf R4.6. Comments by reviewer}:\\
{\em On page 5, is there a mistake in ``random samples from (28), and evaluated at (28)"?}

{\nib \blue{That was not a typo. Indeed, equation (28) was \ul{used in two ways}: it was used to generate initial state samples, and then the initial joint PDF values at those samples were evaluated via the functional form of (28), i.e., as in function evaluation. This is needed in our case because unlike standard Monte Carlo, we \ul{co-evolve} the state samples and joint PDF values at those state samples. \ul{To clarify this further, in the revised manuscript, we have slightly expanded that sentence}.}}

\noindent
{\bf R4.7. Comments by reviewer}:\\
{\em As a validation, it would be good to also provide information of higher-order moments along with the mean given in Fig. 10.}

{\nib \blue{\ul{In the revised manuscript, we have incorporated new Fig. 13 and Fig. 14}. Fig. 13 shows the relative error between the covariances of the Monte Carlo and proximal joint state PDFs, as measured w.r.t. the normalized Bures-Wasserstein distance. \ul{We have explained the same in text in the last but one paragraph of Sec. V-B (highlighted in blue)}. Fig. 14 shows the absolute error between the Monte Carlo and proximal joint state PDFs, as measured w.r.t. the 2-Wasserstein metric. \ul{We have explained the same in text in the last paragraph of Sec. V-B (highlighted in blue, with added inline citations)}.}}

### 2022-03-09 Papers/Stochastic_Uncertainty_Propagation_TPS/JKO_Power_Systems_TPS_Rev1.tex
polishing rebuttal round 1

At each fixed physical time $t_{k}=kh$, $k\in\mathbb{N}$, where $h$ is the fixed step-size, we perform the proximal update $\varrho_{k-1}\mapsto\varrho_{k}$ via Algorithm 1. This proximal update has no analytical solution and instead requires execution of a contractive fixed point recursion. The physical time $t_{k}$ is ``frozen" (zero-order-hold) during the fixed point recursions underlying this one-step proximal update. To claim that the proposed method is practical, we therefore, need to demonstrate that the computational timescale is faster than the physical timescale, i.e., computational times needed to complete the zero-order-hold fixed point recursions are orders-of-magnitude faster than $h$, the step-size of physical time. On the other hand, theoretical consistency (see the last paragraph of Sec. \ref{subsec:ProxUpdate}) requires that $h$ itself should be small. Thus, even if one has a provably convergent algorithm that can perform the one step proximal update, it is not clear \emph{a priori} if the computational time needed to do so can be much faster than $h$. If not, then the method would be statistically consistent but practically not useful. The Fig \ref{fig:CompTimeIEEE14bus} shows that the computational time needed to perform the proximal updates, for all $t_{k}$, are indeed orders-of-magnitude faster than $h$.
}}

@@

Fig. \ref{fig:CompTimeSynthetic} highlights that for computing high dimensional transient joint state PDFs as in here, the proposed variational framework enjoys remarkably fast computational time. Notice that constructing histograms or other conventional density estimators for the joint state PDF by making a grid over the 100 dimensional state space, is computationally prohibitive in a general purpose computer.

>>>>

At each fixed physical time $t_{k}=kh$, $k\in\mathbb{N}$, where $h$ is the fixed step-size, we perform the proximal update $\varrho_{k-1}\mapsto\varrho_{k}$ via Algorithm 1. This proximal update has no analytical solution and instead requires execution of a contractive fixed point recursion. The physical time $t_{k}$ is ``frozen" (zero-order-hold) during the fixed point recursions underlying this one-step proximal update. For the proposed method to be practical, we therefore, need to demonstrate that the computational timescale is faster than the physical dynamics timescale, i.e., computational times needed to complete the zero-order-hold fixed point recursions are orders-of-magnitude faster than $h$, the step-size of physical time. On the other hand, theoretical consistency (see the last paragraph of Sec. \ref{subsec:ProxUpdate}) requires that $h$ itself should be small. Thus, even if one has a provably convergent algorithm that can perform the one step proximal update, it is not clear \emph{a priori} if the computational time needed to do so can be much faster than $h$. If not, then the method would be statistically consistent but practically not useful. The Fig \ref{fig:CompTimeIEEE14bus} shows that the computational time needed to perform the proximal updates, for all $t_{k}$, are indeed orders-of-magnitude faster than $h$.

Fig. \ref{fig:TotalCompTime} (circle markers for the \textbf{Case I}, diamond markers for the \textbf{Case II}) shows the total computational times needed for propagating the transient joint state PDFs over a longer horizon $[0, 1\;\text{min}]$ for 100 different simulation instances.
}}

\begin{figure}[htpb]
\centering
\includegraphics[width=0.95\linewidth]{TotalCompTimeForSimulating1MinHist.png}
\caption{\small{The total computational times for propagating the transient joint state PDFs over $[0, 1\;\text{min}]$ for the simulation set up in Sec. \ref{subsec:NumSimIEEEBus} and Sec. \ref{subsec:NumSimSynthetic}. In each case, we show the total computational time for 100 different simulation instances.}}
\vspace*{-0.1in}
\label{fig:TotalCompTime}
\end{figure}

@@

Fig. \ref{fig:CompTimeSynthetic} highlights that for computing high dimensional transient joint state PDFs as in here, the proposed variational framework enjoys remarkably fast computational time. Notice that constructing histograms or other conventional density estimators for the joint state PDF by making a grid over the 100 dimensional state space, is computationally prohibitive in a general purpose computer. {\blue{Fig. \ref{fig:TotalCompTime} (square markers) shows the total computational times needed for propagating the transient joint state PDFs over the time horizon $[0, 1\;\text{min}]$ for 100 different simulation instances.}}
